Critical assessment of Linear-Response Excitations in Molecular system\\ or \\ On the Localization of the Linear-Response excitations in Open Systems

### 2017-12-20 LRDraft.tex
update

% special operators
\newcommand{\dmnot}{\op{\rho}_0}
\newcommand{\dm}{\op{\rho}}
\newcommand{\hnot}{\op{H}_0}
\newcommand{\hone}[1]{\op{H}_1\argument{#1}}
\newcommand{\transition}[1]{\op T_{#1}}
\newcommand{\excite}[2]{\op e_{{#1}{#2}}}
\newcommand{\decay}[2]{\op d_{{#1}{#2}}}
\newcommand{\honetmp}[1][]{\op{H_1}\argument{#1}} 
\newcommand{\identity}{\op{\mathbb I}}
\newcommand{\rmat}[1]{\optr R}

@@

We introduce the \emph{density operator} defined as the sum of the one-particle densities of all the states with energy lower than the \emph{chemical potential} 
(Fermi energy):
\be\label{rhodef1}
\dmnot = \sum_{\{p\}} \ket{\psi_p} \bra{\psi_p}  \;,
\ee
where the index $p$ labels the occupied state, \emph{i.e} the subset of the bound eigenstates of $\hnot$ with energy lower than the Fermi energy.
Clearly $\dmnot$ commutes with $\hnot$ and in the unperturbed case is a time-independent operator. When the perturbation is switched the system starts to evolve
and we indicate with $\dm(t)$ the modified density of the perturbed system. We introduce the shorthand notation:
\be\lb{rhodef2}
\dm(t)  =  \dmnot + \dm'(t)\;,
\ee
where $\dm'(t)$ is called the \emph{response density operator} and represents the time dependent modification of the ground state density induced by the perturbation.
Here we are assuming that the highest occupied energy level and the lowest unoccupied one will never intersect during the time evolution. Thanks to this condition, the 
linear response regime the expression of $\dm'(t)$ can be readily derived using equations \eqref{ketBraTimeEvol1} and reads:  
\be\lb{rhoTimeEvolutionDef3}
\dm'(t) = \ii \commutator{\dmnot}{\op A(t)} \;.
\ee
The equation of motion (Liouville equation) for $\dm'$ is obtained by performing the time derivative of the previous equation. The derivative of $\op A(t)$ can be 
computed starting from its definition and reads:
\be
\d_t\op A(t) = \op\Phi(t) -\\i \commutator{\hnot}{\op A(t)} \nn \;,
\ee
and using the Jacobi identity we have:
\be\lb{LiouvilleRhoDef1}
\ii \d_t \dm'(t) = \commutator{\op\Phi(t)}{\dmnot} + \ii\commutator{\hnot}{\commutator{\dmnot}{\op A(t)}} = 
\commutator{\op\Phi(t)}{\dmnot} + \commutator{\hnot}{\dm'(t)} \;.
\ee
It is useful to introduce the \emph{Liouvillian} superoperator:
\be\lb{L0sopDef1}
\sop{\mathcal L_0} \equiv \commutator{\hnot}{\cdot} \;,
\ee
so that the Liouville equation can be written as:
\be\lb{LiouvilleRhoDef2}
\ii \d_t \dm'(t) = \sop{\mathcal L_0} \dm'(t) + \commutator{\op\Phi(t)}{\dmnot} \;.
\ee

@@

Let us extend the the former analysis to self-consistent systems, \emph{i.e.} to the case in which the unperturbed Hamiltonian is itself a function of the density.
In this case the perturbing field $\op\Phi$ has also the effect of adding a further perturbation operator to the perturbing Hamiltonian. In the linear response framework, 
this latter quantity can be expressed as the functional derivatives of $\hnot$ with respect to the density kernel (evaluated at $\dm=\dmnot$) times the value of the
induced density modification, that is:
\be\label{V'tspacedef1}
\op V'(t)= 
\int \dd \r \dd \r' \frac{\delta\hnot}{\delta \rho_0(\r,\r')}
\trace{\dm'(t) \ket{\r} \bra{\r'}} \;.
\ee
The Liouville equation for self-consistent systems can be written by adding the contribution of $\op V'$ to $\op\Phi$ in \eqref{LiouvilleRhoDef2}. 
It reads:
\be\lb{LiouvilleRhoSelfDef1}
\ii \d_t \dm'(t) = \sop{\mathcal L_0} \dm'(t) + \commutator{\op V'(t)}{\dmnot} + \commutator{\op \Phi(t)}{\dmnot} \;.
\ee
We point out that the response density operator obeys a self-consistent equation, since the  induced perturbing field $\op V'$ is itself a function of $\dm'$. 
It is possible to formally recast \eqref{LiouvilleRhoSelfDef1} in the same form of \eqref{LiouvilleRhoDef2} by introducing the \emph{coupling}
superoperator:
\be
\sop{\mathcal K} = 
\int \dd \r \dd \r' \commutator{\frac{\delta\hnot}{\delta \rho_0(\r,\r')}}{\dmnot}
\trace{\cdot \ket{\r} \bra{\r'}} \;,
\ee
in this way the first commutator in \eqref{LiouvilleRhoSelfDef1} can be written as the formal action of the coupling superoperator to $\dm'$ and \eqref{LiouvilleRhoSelfDef1}
can be written as:
\be\lb{LiouvilleRhoSelfDef2}
\ii \d_t \dm'(t) = \sop{\mathcal L} \dm'(t) + \commutator{\op \Phi(t)}{\dmnot} \;, \qq
\sop{\mathcal L} = \sop{\mathcal L_0} + \sop{\mathcal K} \;.
\ee

@@

We reformulate the Liouville equation \eqref{LiouvilleRhoSelfDef2} in the frequency domain. In order to achieve this task it is
preferable to directly compute the action of the Liouville operator on the Fourier transform of the response density, since in this
way some subtleties that concerns the analytic structure in the $\omega$ complex plane can be properly taken into account.
We first define the response density in the frequency domain that, staring from \eqref{rhoTimeEvolutionDef3}, is given by:
\be\lb{rhoPrimeOmegaDef1}
\dm'(\omega) \equiv \intover{t} e^{\ii \omega t} \dm'(t)
= \ii \commutator{\dmnot}{\op A(\omega)} \;,
\ee
where $\op A(\omega)$ is The Fourier transform of \eqref{AopDef1} and reads:
\be \lb{AomegaDef1}
\op A(\omega) = \frac{1}{2\pi} \int_{-\infty}^\infty  \dd \omega'
\op G_0^+(\omega') \op \Phi(\omega) \op G_0^-(\omega'-\omega)\;.
\ee
The action of $\sop{\mathcal L_0}$ on $\dm'(\omega)$ follows from \eqref{L0sopDef1}:
\be\lb{LiouvillianRho'omegaDef1}
\sop{\mathcal L_0} \dm'(\omega) = \ii \commutator{\hnot}{\commutator{\dmnot}{\op A(\omega)}} = 
\ii \commutator{\dmnot}{\commutator{\hnot}{\op A(\omega)}} \;,
\ee
and the commutator between the unperturbed Hamiltonian and $\op A(\omega)$ can be directly computed starting from \eqref{AomegaDef1}: 
\be
\commutator{\hnot }{\op A(\omega)} = \bbomega \op A(\omega) - \ii \op \Phi(\omega)\;, \nn
\ee
where $\bbomega \equiv \omega + 2\ii\eta$, so that:
\be\lb{LiouvillianRho'omegaDef2}
\sop{\mathcal L_0} \dm'(\omega) =  \bbomega\dm'(\omega) + \commutator{\dmnot}{\op\Phi(\omega)}\;.
\ee
The coupling superoperator has the same formal action on the response density both in the time and in the frequency domain, so the complete
Liouville equation in $\omega$ space can be written as:
\be\lb{LiouvillianRho'omegaDef3}
\left(\bbomega - \sop{\mathcal L}\right) \dm'(\omega) =  \commutator{\op\Phi(\omega)}{\dmnot}\;.
\ee

@@

We split the complete
set of eigenstates of $\hnot$ in two sets. The first one is composed by the eigenvectors of $\hnot$ in the subspace of the occupied orbitals. The second 
one is composed by the remaining eigenvectors, both in the bound and in the continuum region, and its elements are denoted as $\ket{\psi(\alpha)}$, where $\alpha$ 
is a multi-index in the sense described in \ref{BasicDef}. The elements of these sets satisfy the following orthonormality relation: 
\be\lb{orthonormalityDef1}
\brket{\psi(\alpha)}{\psi_p} = \brket{\psi_p}{\psi(\alpha)} = 0 \;, \qq \brket{\psi_p}{\psi_q} = \delta_{pq} \;, \qq
\brket{\psi(\alpha)}{\psi(\beta)} = \delta(\alpha-\beta) \;.
\ee
Obviously, the direct sum of these two sets coincides with the complete set of eigenstates of $\hnot$ and thus 
constitute a complete basis. 
A generic operator can be decomposed in this basis and its formal expansion contains four distinct contribution of the form $\ketbra{\psi_p}{\psi_q}$, 
$\ketbra{\psi(\alpha)}{\psi(\beta)}$, $\ketbra{\psi(\alpha)}{\psi_p}$, $\ketbra{\psi_p}{\psi(\alpha)}$. However an inspection of equations \eqref{rhoPrimeOmegaDef1}
and \eqref{LiouvillianRho'omegaDef3} evidences that the relevant operators in the contest of the linear response are expressed as a commutator with $\dmnot$. 
For this class of operators the only non vanishing contribution comes from the \emph{transition operators}:
\be
\excite{p}{\alpha} \equiv \ketbra{\psi(\alpha)}{\psi_p}\;, \qq
\decay{\alpha}{p} \equiv \ketbra{\psi_p}{\psi(\alpha)} \;.
\ee
The orthonormality conditions \eqref{orthonormalityDef1} imply that:
\be\lb{transitionOpOrthonormalityDef1}
\trace{\excite{p}{\alpha}\excite{q}{\beta}} = \trace{\decay{\alpha}{p}\decay{\beta}{q}} = 0 \;, \qq
\trace{\excite{p}{\alpha}\decay{\beta}{q}} = \delta_{pq}\delta(\alpha-\beta) \;,
\ee
so that the expansion of a suitable operator $\op O$ in the basis of transitions reads:
\be\lb{OperatorTransBasisExpDef1}
\op O = \sum_p\sint \dd \alpha\left(
\excite{p}{\alpha}\trace{\decay{\alpha}{p}\op O} + \decay{\alpha}{p}\trace{\excite{p}{\alpha}\op O}
\right)
\ee
In order to recast this expansion in a simpler expression it turns out to be useful to define the rank-2 vectors of the transition basis: 
\be
\sbra{p\alpha}{\cdot} \equiv 
\mat{ \excite{p}{\alpha} \;, & \decay{\alpha}{p} }\;,
\ee
as well as a different representation for the trace between the operators and the above mentioned transitions:
\be
\sket{p\alpha}{\op O}  \equiv \mat{ \trace{ \op O \decay{\alpha}{p}} \\ 
\trace{\op O \excite{p}{\alpha}}} \; .
\ee
Thanks to this formalism we have that equation \eqref{OperatorTransBasisExpDef1} can be expressed as:
\be\lb{OperatorTransBasisExpDef2}
\op O = \sum_p\sint \dd \alpha \sbra{p\alpha}{\cdot}\sket{p\alpha}{\op O} = 
\dbraket{\cdot}{\op O} \;,
\ee
where the last equality implicitly define the index-less ``big-shaped'' vectors and the sum over unwritten indexes is understood.

@@

>>>>

% special operators
\newcommand{\dmnot}{\op{\rho}_0}
\newcommand{\dm}{\op{\rho}}
\newcommand{\hnot}{\op{H}_0}
\newcommand{\hone}[1]{\op{H}_1\argument{#1}}
\newcommand{\transition}[1]{\op T_{#1}}
\newcommand{\excite}[2]{\op e_{{#1}{#2}}}
\newcommand{\decay}[2]{\op d_{{#1}{#2}}}
\newcommand{\Liouv}{\sop{\mathcal L}}
\newcommand{\Liouvnot}{\sop{\mathcal L_0}}
\newcommand{\coupl}{\sop{\mathcal K}}
\newcommand{\honetmp}[1][]{\op{H_1}\argument{#1}} 
\newcommand{\identity}{\op{\mathbb I}}
\newcommand{\rmat}[1]{\optr R}

@@

We introduce the \emph{density operator} defined as the sum of the one-particle densities of all the states with energy lower than the \emph{chemical potential} 
(Fermi energy):
\be\label{rhodef1}
\dmnot = \sum_{\{p\}} \ket{\psi_p} \bra{\psi_p}  \;,
\ee
where the index $p$ labels the occupied state, \emph{i.e} the subset of the bound eigenstates of $\hnot$ with energy lower than the Fermi energy.
Clearly $\dmnot$ commutes with $\hnot$ and in the unperturbed case is a time-independent operator. When the perturbation is switched the system starts to evolve
and we indicate with $\dm(t)$ the modified density of the perturbed system. We introduce the shorthand notation:
\be\lb{rhodef2}
\dm(t)  =  \dmnot + \dm'(t)\;,
\ee
where $\dm'(t)$ is called the \emph{response density operator} and represents the time dependent modification of the ground state density induced by the perturbation.
Here we are assuming that the highest occupied energy level and the lowest unoccupied one will never intersect during the time evolution. Thanks to this condition, the 
linear response regime the expression of $\dm'(t)$ can be readily derived using equations \eqref{ketBraTimeEvol1} and reads:  
\be\lb{rhoTimeEvolutionDef3}
\dm'(t) = \ii \commutator{\dmnot}{\op A(t)} \;.
\ee
The equation of motion (Liouville equation) for $\dm'$ is obtained by performing the time derivative of the previous equation. The derivative of $\op A(t)$ can be 
computed starting from its definition and reads:
\be
\d_t\op A(t) = \op\Phi(t) -\\i \commutator{\hnot}{\op A(t)} \nn \;,
\ee
and using the Jacobi identity we have:
\be\lb{LiouvilleRhoDef1}
\ii \d_t \dm'(t) = \commutator{\op\Phi(t)}{\dmnot} + \ii\commutator{\hnot}{\commutator{\dmnot}{\op A(t)}} = 
\commutator{\op\Phi(t)}{\dmnot} + \commutator{\hnot}{\dm'(t)} \;.
\ee
It is useful to introduce the \emph{Liouvillian} superoperator:
\be\lb{L0sopDef1}
\Liouvnot \equiv \commutator{\hnot}{\cdot} \;,
\ee
so that the Liouville equation can be written as:
\be\lb{LiouvilleRhoDef2}
\ii \d_t \dm'(t) = \Liouvnot \dm'(t) + \commutator{\op\Phi(t)}{\dmnot} \;.
\ee

@@

Let us extend the the former analysis to self-consistent systems, \emph{i.e.} to the case in which the unperturbed Hamiltonian is itself a function of the density.
In this case the perturbing field $\op\Phi$ has also the effect of adding a further perturbation operator to the perturbing Hamiltonian. In the linear response framework, 
this latter quantity can be expressed as the functional derivatives of $\hnot$ with respect to the density kernel (evaluated at $\dm=\dmnot$) times the value of the
induced density modification, that is:
\be\label{V'tspacedef1}
\op V'(t)= 
\int \dd \r \dd \r' \frac{\delta\hnot}{\delta \rho_0(\r,\r')}
\trace{\dm'(t) \ket{\r} \bra{\r'}} \;.
\ee
The Liouville equation for self-consistent systems can be written by adding the contribution of $\op V'$ to $\op\Phi$ in \eqref{LiouvilleRhoDef2}. 
It reads:
\be\lb{LiouvilleRhoSelfDef1}
\ii \d_t \dm'(t) = \Liouvnot \dm'(t) + \commutator{\op V'(t)}{\dmnot} + \commutator{\op \Phi(t)}{\dmnot} \;.
\ee
We point out that the response density operator obeys a self-consistent equation, since the  induced perturbing field $\op V'$ is itself a function of $\dm'$. 
It is possible to formally recast \eqref{LiouvilleRhoSelfDef1} in the same form of \eqref{LiouvilleRhoDef2} by introducing the \emph{coupling}
superoperator:
\be
\coupl = 
\int \dd \r \dd \r' \commutator{\frac{\delta\hnot}{\delta \rho_0(\r,\r')}}{\dmnot}
\trace{\cdot \ket{\r} \bra{\r'}} \;,
\ee
in this way the first commutator in \eqref{LiouvilleRhoSelfDef1} can be written as the formal action of the coupling superoperator to $\dm'$ and \eqref{LiouvilleRhoSelfDef1}
can be written as:
\be\lb{LiouvilleRhoSelfDef2}
\ii \d_t \dm'(t) = \Liouv \dm'(t) + \commutator{\op \Phi(t)}{\dmnot} \;, \qq
\Liouv = \Liouvnot + \coupl \;.
\ee

@@

We reformulate the Liouville equation \eqref{LiouvilleRhoSelfDef2} in the frequency domain. In order to achieve this task it is
preferable to directly compute the action of the Liouville operator on the Fourier transform of the response density, since in this
way some subtleties that concerns the analytic structure in the $\omega$ complex plane can be properly taken into account.
We first define the response density in the frequency domain that, staring from \eqref{rhoTimeEvolutionDef3}, is given by:
\be\lb{rhoPrimeOmegaDef1}
\dm'(\omega) \equiv \intover{t} e^{\ii \omega t} \dm'(t)
= \ii \commutator{\dmnot}{\op A(\omega)} \;,
\ee
where $\op A(\omega)$ is The Fourier transform of \eqref{AopDef1} and reads:
\be \lb{AomegaDef1}
\op A(\omega) = \frac{1}{2\pi} \int_{-\infty}^\infty  \dd \omega'
\op G_0^+(\omega') \op \Phi(\omega) \op G_0^-(\omega'-\omega)\;.
\ee
The action of $\Liouvnot$ on $\dm'(\omega)$ follows from \eqref{L0sopDef1}:
\be\lb{LiouvillianRho'omegaDef1}
\Liouvnot \dm'(\omega) = \ii \commutator{\hnot}{\commutator{\dmnot}{\op A(\omega)}} = 
\ii \commutator{\dmnot}{\commutator{\hnot}{\op A(\omega)}} \;,
\ee
and the commutator between the unperturbed Hamiltonian and $\op A(\omega)$ can be directly computed starting from \eqref{AomegaDef1}: 
\be
\commutator{\hnot }{\op A(\omega)} = \bbomega \op A(\omega) - \ii \op \Phi(\omega)\;, \nn
\ee
where $\bbomega \equiv \omega + 2\ii\eta$, so that:
\be\lb{LiouvillianRho'omegaDef2}
\Liouvnot \dm'(\omega) =  \bbomega\dm'(\omega) + \commutator{\dmnot}{\op\Phi(\omega)}\;.
\ee
The coupling superoperator has the same formal action on the response density both in the time and in the frequency domain, so the complete
Liouville equation in $\omega$ space can be written as:
\be\lb{LiouvillianRho'omegaDef3}
\left(\bbomega - \Liouv\right) \dm'(\omega) =  \commutator{\op\Phi(\omega)}{\dmnot}\;.
\ee

@@

We split the complete
set of eigenstates of $\hnot$ in two sets. The first one is composed by the eigenvectors of $\hnot$ in the subspace of the occupied orbitals. The second 
one is composed by the remaining eigenvectors, both in the bound and in the continuum region, and its elements are denoted as $\ket{\psi(\alpha)}$, where $\alpha$ 
is a multi-index in the sense described in \ref{BasicDef}. The elements of these sets satisfy the following orthonormality relation: 
\be\lb{orthonormalityDef1}
\brket{\psi(\alpha)}{\psi_p} = \brket{\psi_p}{\psi(\alpha)} = 0 \;, \qq \brket{\psi_p}{\psi_q} = \delta_{pq} \;, \qq
\brket{\psi(\alpha)}{\psi(\beta)} = \delta(\alpha-\beta) \;.
\ee
Obviously, the direct sum of these two sets coincides with the complete set of eigenstates of $\hnot$ and thus 
constitute a complete basis. 
A generic operator can be decomposed in this basis and its formal expansion contains four distinct contribution of the form $\ketbra{\psi_p}{\psi_q}$, 
$\ketbra{\psi(\alpha)}{\psi(\beta)}$, $\ketbra{\psi(\alpha)}{\psi_p}$, $\ketbra{\psi_p}{\psi(\alpha)}$. However an inspection of equations \eqref{rhoPrimeOmegaDef1}
and \eqref{LiouvillianRho'omegaDef3} evidences that the relevant operators in the contest of the linear response are expressed as a commutator with $\dmnot$. 
For this class of operators the only non vanishing contribution comes from the \emph{transition operators}:
\be
\excite{p}{\alpha} \equiv \ketbra{\psi(\alpha)}{\psi_p}\;, \qq
\decay{\alpha}{p} \equiv \ketbra{\psi_p}{\psi(\alpha)} \;.
\ee
The orthonormality conditions \eqref{orthonormalityDef1} imply that:
\be\lb{transitionOpOrthonormalityDef1}
\trace{\excite{p}{\alpha}\excite{q}{\beta}} = \trace{\decay{\alpha}{p}\decay{\beta}{q}} = 0 \;, \qq
\trace{\excite{p}{\alpha}\decay{\beta}{q}} = \delta_{pq}\delta(\alpha-\beta) \;,
\ee
so that the expansion of a suitable operator $\op O$ in the basis of transitions reads:
\be\lb{OperatorTransBasisExpDef1}
\op O = \sum_p\sint \dd \alpha\left(
\excite{p}{\alpha}\trace{\decay{\alpha}{p}\op O} + \decay{\alpha}{p}\trace{\excite{p}{\alpha}\op O}
\right)
\ee
In order to recast this expansion in a simpler expression it turns out to be useful to define the rank-2 vectors of the transition basis: 
\be
\sbra{p\alpha}{\cdot} \equiv 
\mat{ \excite{p}{\alpha} \;, & \decay{\alpha}{p} }\;,
\ee
as well as a different representation for the trace between the operators and the above mentioned transitions:
\be\lb{OperatorTransitionCoeffDef1}
\sket{p\alpha}{\op O}  \equiv \mat{ \trace{\decay{\alpha}{p}\op O } \\ 
\trace{\excite{p}{\alpha}\op O}} \; .
\ee
Thanks to this formalism we have that equation \eqref{OperatorTransBasisExpDef1} can be expressed as:
\be\lb{OperatorTransBasisExpDef2}
\op O = \sum_p\sint \dd \alpha \sbra{p\alpha}{\cdot}\sket{p\alpha}{\op O} = 
\dbraket{\cdot}{\op O} \;,
\ee
where the last equality implicitly define the index-less ``big-shaped'' vectors and the sum over unwritten indexes is understood.

@@

The Liouville equation \eqref{LiouvillianRho'omegaDef3} can be written in the basis of transititions and becomes a linear matrix equation for the rank-2 vectors
of the coefficients \eqref{OperatorTransitionCoeffDef1} of the operators in its left and right side:
\be
\omega\mat{ \trace{\decay{\alpha}{p}\dm'(\omega)} \\ \trace{\excite{p}{\alpha}\dm'(\omega)}} - 
\mat{ \trace{\decay{\alpha}{p}\Liouv\dm'(\omega)} \\ \trace{\excite{p}{\alpha}\Liouv\dm'(\omega)}} =
\mat{ \trace{\decay{\alpha}{p}\commutator{\op\Phi(\omega)}{\dmnot}} \\ \trace{\excite{p}{\alpha}\commutator{\op\Phi(\omega)}{\dmnot}}}
\;.\nn
\ee
The second addend in the l.h.s can be further analyzed by writing the transition basis expansion of $\dm'$ inside the traces. Performing this procedure
we can recast this quantity as the action of a 2-dimensional matrix on the rank-2 vector of the response density. So we have: 
\be\lb{LiouvilleTransBasisDef1}
\sum_q\sint\dd\beta \left[\omega\delta_q^p\delta(\alpha-\beta)- \optr{L}^{p \alpha}_{q\beta} \right]\sket{q\beta}{\dm'(\omega)} =
\sket{p\alpha}{\commutator{\op\Phi(\omega)}{\dmnot}} \;.
\ee
with:
\be
\optr{L}^{p \alpha}_{q\beta} \equiv \mat{\trace{\decay{\alpha}{p}\, \Liouv\, \excite{q}{\beta}} &
\trace{\decay{\alpha}{p}\, \Liouv\, \decay{\beta}{q}} \\
\trace{\excite{p}{\alpha}\, \Liouv\, \excite{q}{\beta}} &
\trace{\excite{p}{\alpha}\, \Liouv\, \decay{\beta}{q}}} \;.
\ee
